Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-08-20 manuscript.qmd
describe llh in Bayes' context for bilateral
Related: #11

Our stated goal is to compute the risk for a patient's true nodal involvement state $\mathbf{X}$, *given* their individual diagnosis $d = \left( \boldsymbol{\zeta}_k, \text{T}x \right)$. Using Bayes' law, this is written as:

$$
P \big( \mathbf{X} \mid \mathbf{Z}=\boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \frac{P \left( \boldsymbol{\zeta}_k \mid \mathbf{X} \right) P \big( \mathbf{X} \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}{\sum_{i=0}^{2^V} P \left( \boldsymbol{\zeta}_k \mid \mathbf{X}=\boldsymbol{\xi}_i \right) P \big( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
$$ {#eq-uni-bayes-law}

The term $P \left( \boldsymbol{\zeta}_k \mid \mathbf{X} \right)$ is defined solely by sensitivity and specificity of the diagnostic modality. Terms like this already appeared in the definition of the observation matrx in @eq-obs-matrix. The *prior* $P \big( \mathbf{X} \mid \boldsymbol{\hat{\theta}} \big)$ in the above equation is the crucial term that is supplied by a trained model and its parameters $\boldsymbol{\hat{\theta}}$.

It is possible to compute this *posterior* probability of true involvement not only for one fully defined state $\mathbf{X}$, but also for e.g. individual LNLs: For example, the risk for involvement in level IV would be a marginalization over all states $\boldsymbol{\xi}_i$, where $\xi_{i4}=1$. Formally:

$$
P \big( \text{IV} \mid \mathbf{Z}=\boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x  \big) = \sum_{k \, : \, \xi_{k4}=1} P \big( \mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x  \big)
$$ {#eq-marg-over-posterior}

@@

>>>>

@@

Our stated goal is to compute the risk for a patient's true ipsi- and contralateral nodal involvement states $\mathbf{X}^\text{i}$ and $\mathbf{X}^\text{c}$, *given* their individual diagnosis $d = \left( \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \text{T}x \right)$. Here, this diagnosis consists of the observed ipsi- and contralateral nodal involvements, the patient's midline extension $\epsilon$, and their tumor's T-category $\text{T}x$. Using Bayes' law, we can write this risk as:

$$
P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid d, \boldsymbol{\hat{\theta}} \big)
= \frac{P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right) P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right) P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
{\sum_{i=0}^{2^V} \sum_{j=0}^{2^V} P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i \right) P \big( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j \big) P \big( \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i, \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
$$ {#eq-uni-bayes-law}

The terms $P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right)$ and $P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right)$ are defined solely by sensitivity and specificity of the diagnostic modality. Terms like these already appeared in the definition of the observation matrx in @eq-obs-matrix. The *prior* $P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)$ in the above equation is the crucial term that is supplied by a trained model and its parameters $\boldsymbol{\hat{\theta}}$.

It is possible to compute this *posterior* probability of true involvement not only for one fully defined state $(\mathbf{X}^\text{i}, \mathbf{X}^\text{c})$, but also for e.g. individual LNLs: For example, the risk for involvement in the contralateral level IV would be a marginalization over all ipsilateral states $\boldsymbol{\xi}^\text{i}_i$ and all contralateral states $\boldsymbol{\xi}^\text{c}_j$ where $\xi^\text{c}_{j4}=1$. Formally:

$$
P \big( \text{cIV} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
$$ {#eq-marg-over-posterior}
